Distribution-oblivious Programming with Dish

### 2019-11-21 pldi/draft.tex
A syntax for the DSL, and some progress on Front end

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{wrapfig}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_schematic.pdf}
\caption{
  \textbf{\sys high-level schematic.}
	\sys XYZ
}
\label{fig:schematic}
\end{figure}

@@

% \begin{figure*}[t]
% \centering
% \includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
% \caption{
%   \textbf{Applying \sys to $p_1$.}
% }
% \label{fig:example}
% \end{figure*}
%

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

@@

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

@@

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{./figs/dish_ex3.pdf}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
As described before, the file can be a FIFO dubbing as input to another command.
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_overview.pdf}
\caption{
  \textbf{\sys transformation overview.}
	\sys XYZ
}
\label{fig:overview}
\end{figure}

@@

@@

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \texttt{Dish} first translates it to
a dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour.

\kk{Give examples of writing to environment variables, producing
  irreversible side effects, the ; and \&\& operators.}

In order to address this issue, we introduce the notion of
distributable AST subtrees, and we design an analysis that given the
shell script Abstract Syntax Tree identifies distributable AST
subtrees and translates each one of them to a dataflow graph. The AST
of the shell script is produced using a POSIX compliant parser
\kk{Reference libdash github}.

\subsection{Distributable Subtrees}

\kk{The goal of this section is to convince the reader ``intuitively''
  that we choose safe parts of a shell program to distribute. There is
  no formal argument, but one should be able to intuitively believe
  us.}

@@

\kk{Give an example of a shell script that cannot be distributed
  completely, Possibly involving ; and writing to a variable.}

\kk{Intuitively explain what could be distributed and what not.}

@@

The analysis is done in a DFS fashion on the AST. Distributable
subtrees grow from the bottom up, until a construct that doesn't
``allow'' them to go through is encountered.

%% Mention all the different constructs that our front end handles, and
%% intuitively mention why it is ``safe'' to add them to the distribution
%% pipeline.

Subtrees don't cross the following:

\begin{itemize}
\item Sequence, And,  or
\item Function definitions
\item Command Substitutions
\item control flow: while, for, if, case: While these constructs can
  be parallelized and there has been research on it, it is orthogonal
  to our work, and can be incorporated as future work.
\end{itemize}

\kk{Maybe mention the recursive case. I am not sure if that should be
  here, or in the distribution planner. In any case, it should be
  after we say that an IR is replaced with a call to the planner with
  the IR saved in a file. The example is a command in a pipeline whose
  first argument is a command substitution with a pipeline. Executing
  the outside dataflow graph, would lead to the command to be executed
  as part of this shell. However its first argument would have a call
  to the planner itself, leading to another distributed implementation
  running in parallel. The rest of the pipeline would wait for this to
  end before it continues. Assuming that the files that both of these
  graphs use are different, and that the above assumptions hold, there
  should be no interference between the two graphs and they should be
  able to run in parallel}

@@

@@

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{./figs/dish_spectrum.pdf}
\caption{
  \textbf{Cost of Manual Effort.}
	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness. A more complete picture is presented in the related-work section~\sx{rt}.
}
\label{fig:spectrum}
\end{figure}

>>>>

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{wrapfig}
\usepackage{syntax}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_schematic.pdf}}
\caption{
  \textbf{\sys high-level schematic.}
	\sys XYZ
}
\label{fig:schematic}
\end{figure}

@@

% \begin{figure*}[t]
% \centering
% \includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
% \caption{
%   \textbf{Applying \sys to $p_1$.}
% }
% \label{fig:example}
% \end{figure*}
%

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{\detokenize{./figs/dish_ex1.pdf}}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

@@

% TODO: Talk about security primitives?
Streams are a key feature of \unix, and go hand-in-hand with the file abstraction:
  files are simply persistent streams that can be manipulated by commands or pipelines.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
Reading a file or a set of files generates a stream; 
  similarly, streams can be trivially redirected to the file system---by appending with the file identifier.
Files are named, in the sense that they correspond to a global identifier---for example \ttt{cat /x/y} returns the persistent stream associated with the identifier \ttt{/x/y}.
\unix exposes a few different types of files---\eg directories, symbolic links, named pipes (or FIFOs), and special character devices.
As a result, file-system identifiers part of a pipeline may resolve to in-memory structures or pseudo-devices (\eg \ttt{/dev/urandom}).

@@

A small but powerful set of features is concerned with stream manipulation.
Some of these features are built into the shell;
  for example, given a file identifier, stream redirection operators can direct stdout or stderr to a file.
\begin{wrapfigure}{r}{0.15\textwidth}
  \vspace{-15pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.15\textwidth]{\detokenize{./figs/dish_ex3.pdf}}
  % \end{center}
  \vspace{-25pt}
\end{wrapfigure}
As described before, the file can be a FIFO dubbing as input to another command.
More interestingly, some commands allow splitting (resp. merging) a single stream (resp. many streams) into multiple streams (resp. a single stream);
  stream merging can be achieved by appending one stream after the end of another or by fusing (zipping) streams column-wise into a single (fat) stream.
% TODO: talk about \ttt{xargs}?

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_overview.pdf}}
\caption{
  \textbf{\sys transformation overview.}
	\sys XYZ
}
\label{fig:overview}
\end{figure}

@@

\kk{New sketch of the language}

We assume that commands satisfy the standard that options and flags
come first, and then file arguments. The language contains a record
for each command name, together with a sequence of predicates that map
to command categories and designation of input output.

%% Can be tricked according to this:
%% https://tex.stackexchange.com/questions/24886/which-package-can-be-used-to-write-bnf-grammars

\begin{grammar}
  <flag> ::= `-' <string>

  <category> ::= `stateless' | `pure' | ...

  <arg> ::= `arg[' <int> `]'

  <input> ::= `stdin' | <arg>

  <output> ::= `stdout' | <arg>

  <flag-pred> ::= <flag>
  \alt `value' <flag> = <string>
  \alt `!' <flag-pred>
  \alt <flag-pred> `or' <flag-pred>
  \alt <flag-pred> `and' <flag-pred>

  <assignment> ::= `(' <category>, <input>, <output> `)'

  <predicate> ::= <flag-pred> `=>' <assignment>

  <pred-list> ::= `|' <predicate> <pred-list>
  \alt `|' `otherwise' `=>' <assignment>

  <command> ::= <name> `:' <pred-list>

\end{grammar}

\kk{We don't need to designate the stderror, because we assume that it
  is never the main output of a command and it will never be used by
  the input of another command in the pipeline. If someone indeed
  wants to do this, they can just use a redirect I think to get around
  it.}

\kk{Give an example of a flag that has value and how this could alter
  category or input etc...}

@@

As mentioned in \Cref{ir}, in order to generate the distributed
implementation of a shell script, \texttt{Dish} first translates it to
a dataflow graph. However, there are several components in shell
programs that cannot be arbitrarily parallelized without affecting the
program behaviour, e.g. commands producing an irreversible effect, or
commands connected with a \texttt{\&\&} operator. In addition, the
fact that the shell is \todo{extremely dynamic}, further increases the
challenge of correctly identifying what parts can be distributed.

%% \kk{Give examples of writing to environment variables, producing
%%   irreversible side effects, the ; and \&\& operators.}

In order to address this issue, we introduce the notion of
distributable program regions, and we design an analysis that given the
abstract syntax tree of a shell script identifies distributable
regions and translates each of them to a dataflow graph. The
AST of the shell script is produced using a POSIX compliant parser
\kk{Reference libdash github. This reference should probably go to
  evaluation}.

\subsection{Distributable Regions}

\kk{I wanted this to be purely about the regions, and a next
  subsection be about the analysis, but the separation line is very
  blurry, so they are now in the same subsection.}

\kk{The goal of this subsection is to explain that Dish does a safe
  but effective analysis to identify which parts of the scripts to
  parallelize.}

@@

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  cat f1 f2 | grep ``foo'' > f3 &&
  sort f3 #Replace
\end{lstlisting}

Notice that the structure of a shell program clearly identifies
components that can be executed concurrently, and barriers that
enforce synchronization. More precisely, \texttt{cat f1 f2} and
\texttt{grep ``foo''} already execute concurrently in the standard
shell, while \texttt{sort f3} waits for their completion before being
executed. Both \texttt{cat f1 f2 | grep ``foo'' > f3} and \texttt{sort
  f3} are maximal distributable regions, in the sense that they
cannot be extended beyond \texttt{\&\&}.

Concretely, a distributable region cannot permeate through the
following shell constructs: ;, \&\&, \texttt{||}, function
definitions, command substitutions, and control flow primitives. On
the other hand, the pipe operator \texttt{|} and the parallel
composition operator \& enable parallelization, and extend
distributable regions. Essentially, distributed regions correspond to
subsections of the program that would be executed by different shell
processes.

\kk{I don't know whether I should mention the following: While these
  control flow constructs can be parallelized and there has been
  research on it, it is orthogonal to our work, and can be
  incorporated as future work.}

Dish performs a depth first search pass on the AST of the given shell
program, extending the distributable regions from the bottom
up, translating their \todo{independent} components to dataflow graph
nodes, until it reaches a barrier construct. Note that barrier
constructs can be contained in a distributable subexpression, as a
\todo{whole independent component}. An illustrative example follows:

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  (sort f1 > g1 && sort f2 > g2) &
  sort -m g1 g2
\end{lstlisting}

In the above example\kk{Can we make example not floating?}, while
\texttt{sort f1 > g1}, \texttt{sort f2 > g2}, and \texttt{sort -m g1
  g2} are all in different distributable regions. The dataflow graph
that corresponds to the \todo{large} region, has two nodes,
\texttt{sort f1 > g1 \&\& sort f2 > g2} and \texttt{sort -m g1 g2}.

The analysis algorithm also finds the category of each node that it
adds to the dataflow graph, using the rules described in \ref{}. Some
interesting cases of the analysis algorithm are shown in pseudocode
below:

\begin{lstlisting}[language=python, float=h]
  def analysis(node):
    ...
    if(node.name == 'Pipe'):
      return pipe_graphs([analysis(child)
        for child in node.children])
    elif(node.name == 'And'):
      node.children = [analysis(child)
        for child in node.children]
      return node
    elif(node.name == 'Command'):
      if(not safe(node)):
        return make_ir_node(node, 'unsafe')
      else:
        return make_ir_node(node,
          find_category(node))
    ...
\end{lstlisting}

\kk{Explain how the three above cases work, mentioning what does
  pipe-graphs and safe do. The important point to get across is to
  explain why we do the ``conservative checks'' in safe, and how they
  preserve safety.}

\paragraph{Assumptions}

\kk{Rephrase that paragraph, to mention the assumtpions that the front
  end makes, and why they are not so unreasonable}

There are some assumptions that we make in order to preserve the
correctness of the computation under distributability optimizations.

@@

@@

\kk{This section could be moved to the intermediate representation,
  leaving a small section about the iterative process of the planner
  in the implementation/runtime section}

@@

\kk{Mention the assumptions that need to hold for the graph
  transformations to be valid. Commands must be deterministic, they
  must not do any other side effect (such as writing to other files,
  sending signals, etc). However, these assumptions must already be
  checked when the developer designates the categories.}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.49\textwidth]{\detokenize{./figs/dish_spectrum.pdf}}
\caption{
  \textbf{Cost of Manual Effort.}
	\sys sits at the automation end of the spectrum, automatically distributing shell pipelines while maintaining their correctness. A more complete picture is presented in the related-work section~\sx{rt}.
}
\label{fig:spectrum}
\end{figure}
